---
Title: "Neural Network Modeling with R"
author: "Lavina Kathambi"
date: "2023-09-21"
output:
  pdf_document: default
  html_document: default
---


# Installing & Loading Relevant Libraries 
```{r}
# install.packages("neuralnet") # Used for building and training neural networks
library(dplyr)
library(neuralnet)

```

# Data Loading and Exploration:
```{r}
mydata <- read.csv("/Users/lavinakathambi/Desktop/R/AI/lab3/Lab3_all_by_stu_term.csv", header = TRUE)
head(mydata)
```
# Display basic statistics of the loaded data
```{r}
summary(mydata)
```

# Data Preprocessing:
```{r}
# Filter out BMI outliers, keeping only values between 0 and 100
mydata <- subset(mydata, bmi > 0 & bmi < 100)

# Drop rows with missing values in 'bmi', 'bmi_gender', and 'fitbit_steps'
mydata <- mydata[complete.cases(mydata$bmi, mydata$bmi_gender, mydata$fitbit_steps), ]

# Convert 'bmi_gender' from M/F to numeric (1/0)
mydata$bmi_gender <- ifelse(mydata$bmi_gender == "M", 1, 0)

head(mydata[c("bmi", "bmi_gender", "fitbit_steps")])
```

# Data Splitting:
```{r}

# Split the data into a training set (80%) and a test set (20%)
set.seed(123) # For reproducibility 
data_rows <- floor(0.80 * nrow(mydata)) # Calculating the Number of Rows for Training = 80% of data
train_indices <- sample(c(1:nrow(mydata)), data_rows) #Creating Random Training Indices = 80% of data size
train_data <- mydata[train_indices,] # Subsetting the training data from the dataset
test_data <- mydata[-train_indices,] # Subsetting the testing data from the dataset

head(train_indices)
head(train_data)
head(test_data)

```

# Neural Network Modeling:

```{r}
# Train a neural network model using the specified architecture
model = neuralnet(
  term_gpa ~ bmi + bmi_gender + fitbit_steps,
  data=train_data,
  hidden=c(4,2), # Defining our hidden layers
  linear.output = TRUE #TRUE = regression task
)

```

# Visualizing the trained neural network model
```{r}
plot(model,rep = "best")
```

# Model Evaluation:

```{r}
# Generating predictions on the test set using the trained model
pred <- predict(model, test_data)

# Computing the residuals between the predicted and actual values
residuals <- test_data$term_gpa - pred

# Calculating total variance of the actual values
TSS <- sum((test_data$term_gpa - mean(train_data$term_gpa))^2)
TSS

# Compute the variance of the residuals
RSS <- sum(residuals^2)
RSS

# Determine the proportion of variance explained by the model (R^2)
R2 <- 1 - (RSS/TSS)
print(R2)

# Calculate the average prediction error (RMSE)
RMSE <- sqrt(mean(residuals^2))
print(RMSE)
```
In summary, while the model has some explanatory power, it does not explain a large proportion of the variance in term_gpa, and the average prediction error (RMSE) is approximately 0.7547 units. Further analysis and potentially model improvement may be needed depending on the specific goals and requirements of your analysis or prediction task.

